\textsl{Position:} Is machine learning good or bad for the natural sciences?

### 2024-05-31 paper/main.tex
new camera-ready version, taking some comments

% Hogg typesetting issues that won't conflict with ICML
\frenchspacing
\newcommand{\documentname}{\textsl{Position Paper}}
\renewcommand{\paragraph}[1]{\noindent\par\textbf{#1}}
\newcommand{\sectionname}{Section}
\newcommand{\secref}[1]{\sectionname~\ref{#1}}
\newcommand{\figref}[1]{\figurename~\ref{#1}}
\newenvironment{hoggnumerate}
  {\begin{list}
    {\arabic{enumii}.}
    {\usecounter{enumii}
     %\setlength{\labelwidth}{3em}
     %\setlength{\labelsep}{0em}
     \setlength{\itemsep}{0in}
     \setlength{\parsep}{0in}
     \setlength{\parskip}{0in}
     %\setlength{\leftmargin}{1.5cm}
     %\setlength{\rightmargin}{2cm}
     %\setlength{\itemindent}{0em} 
     %\let\makelabel=\makeboxlabel
    }
  }
{\end{list}}

@@

\printAffiliationsAndNotice{}
\begin{abstract}
  Machine learning (ML) methods are having a huge impact across all of the sciences.
  However, ML has a strong ontology---in which only the data exist---and a strong epistemology---in which a model is considered good if it performs well on held-out training data.
  These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
  Here, we identify some locations for ML in the natural sciences at which the ontology and epistemology are valuable.
  For example, when an expressive machine learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can make the results more trustworthy.
  We also show that there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
  For one, when ML models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
  For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint or ensemble analyses without taking on uncontrolled biases.
  The question in the title is being asked of all of the natural sciences; that is, we are calling on the scientific communities to take a step back and consider the role and value of ML in their fields; the (partial) answers we give here come from the particular perspective of physics.
\end{abstract}
% ] % For the arXiv version, uncomment this square bracket

@@

This definition is broad:
In addition to methods like convolutional neural networks \cite{cnn}, multi-layer perceptrons, and transformers \cite{transformer}, it includes large linear regressions, gaussian processes \cite{gp}, support-vector machines \cite{svm}, principal components analysis \cite{pca}, kernel density estimates \cite{kde}, and even some kinds of multilayer or hierarchical models \cite{multilevel}; indeed almost any models in any current ML textbooks (eg, \citealt{ml_book1, ml_book2}).
Feel free to personally take any more restrictive definition.
Our comments will apply to everything in the larger class.

@@

In Bayesian language, this problem is closely related to the point that when data are to be combined, they should be combined at the likelihood level, not the posterior level.
Likelihoods are not affected by population-level biases.
Likelihoods can produce unbiased point estimates.
Likelihood-based estimators are not usually the lowest-risk estimators, but they can be low- or zero-bias estimators.
The outputs of regressions, on the other hand, are more like posterior-based estimators, affected by the implicit prior set by the training set.
When they are combined, the implicit prior gets amplified.
Most ML regressions are not capable of generating or emulating likelihoods.
There are interesting approaches in development to replace standard regressions with generative models that can produce likelihoods \cite{cannon, likelihood_ratio, biwei}.

@@

Finally, we note that one objection to this confirmation-bias point is that perhaps it exists just as strongly for any simulation-based inference?
After all, simulations are expensive, so the full suite of them probably can't be affordably repeated.
There are two responses to this objection.
The first is that if the emulators work---that is, if they greatly reduce the cost of the inference---then the emulators have enormously increased the cost difference (or cost factor) between the inference and any re-analysis using physics-based simulations.
The second is that emulators are not doing precisely what the simulations did.
One way to see this is that they are subject to adversarial attacks \cite{melchior}.
Therefore we know that the emulator is not doing precisely what the physics-based simulation is doing.
That is, verification is not just more expensive, but also more important, when an emulator has been introduced.

@@

Given everything we have said here, what are the ways forward for the natural sciences?
ML is critical in many projects; how do we make sure it is doing what we want it to do?
Our view is that the promising directions include imposing physically motivated mathematical structure into the design or regularization of the ML models.
This includes symmetries, conservation laws, and constraints \cite{bronstein2021geometric, villar2023towards}, but also ODE/PDE solvers \cite{karniadakis2021physics}, dynamic programming modules \cite{xu2020can}, and other established mathematical tools.
The research direction that looks towards explainability, interpretability \cite{interpretable}, and trustworthiness \cite{kearns2019ethical} is also fundamental to the development of ML methods for the physical sciences.
These approaches can increase natural scientists’ confidence in ML tools, and also perhaps expand their ontologies and epistemologies.

@@

\paragraph{Acknowledgements:}
It is a pleasure to thank
  Gaby Contardo (SISSA),
  Jennifer Hill (NYU),
  Adrian Price-Whelan (Flatiron),
  Hans-Walter Rix (MPIA),
  Sam Roweis (deceased), and
  Bernhard Sch\"olkopf (MPI-IS)
for conversations over the years related to these arguments, and
  Vedant Chandra (Harvard),
  Contardo,
  Valentino Foit (NYU),
  Alex Jiang (CUNY),
  Charles Margossian (Flatiron),
  Rix,
  Kate Storey-Fisher (DIPC),
  Kaze Wong (Flatiron), and
  our four anonymous ICML reviewers
for valuable comments on this particular manuscript.
SV was partially supported by ONR N00014-22-1-2126, NSF CCF 2212457, the NSF–Simons Research Collaboration
on the Mathematical and Scientific Foundations of Deep Learning (MoDL) (NSF DMS 2031985), and NSF
CAREER 2339682.
The Flatiron Institute is a division of the Simons Foundation.

@@

To illustrate these ideas, we construct a toy data analysis.
This toy will make a general point, but it has a decidedly astronomical feel to it.
The complete toy data set is generated by a process with the algorithm below.
This algorithm makes references to objects $\xi_\eta$ and $\xi_\zeta$ which are generated prior to the start.
\begin{hoggnumerate}
    \item A floating-point value of a known parameter called ``guiding radius'' $r$ is generated in the range $0<r<14$.
    \item A floating-point value of a latent parameter called ``age'' $\eta$ is generated from a Gaussian with mean $14 - r$ and variance 4.
    \item The point is discarded if the age is outside the range $0<\eta<14$. If the point is not discarded:
    \item A $K$-dimensional latent vector $\zeta$ with $K=14$ is generated from a Gaussian with zero mean and variance 1.
    \item A $M$-dimensional latent vector $\xi$ with $M=110$ is created by $\xi = \xi_\eta\,\eta + \xi_\zeta\cdot\zeta$, where $\xi_\eta$ is a $M$-vector and $\xi_\zeta$ is a $M\times K$ matrix.
    The elements of $\xi_\eta$ and $\xi_\zeta$ were drawn (before the start) from Gaussians with zero mean and variances $0.01/K$ and $1/K$ respectively.
    \item A rectified latent vector $\tilde{\xi}$ is created by taking $1 + \xi$ and rectifying all pixels to lie in the range $[0, 1]$.
    \item A $M$-dimensional feature vector $x$ called ``the data'' is generated from a Gaussian with mean $\tilde{\xi}$ and variance $0.0025$. (These feature vectors have been designed to look a little like stellar spectra.)
    \item A label $y$ called ``measured age'' is generated from a Gaussian with mean $\eta$ and variance 1.
\end{hoggnumerate}
The data generated this way are shown in the top panels of \figref{fig:regression}.
The data contain a non-trivial relationship between label $y$ (measured age) $y$ and known parameter $r$ (guiding radius).
This relationship is shown in \figref{fig:regression}.
It isn't precisely linear because of the removal of data points with latent ages $\eta$ outside the range $0<\eta<14$.

>>>>

% Hogg typesetting issues that won't conflict with ICML
\raggedbottom\frenchspacing
\newcommand{\documentname}{\textsl{Position Paper}}
\renewcommand{\paragraph}[1]{\noindent\par\textbf{#1}}
\newcommand{\sectionname}{Section}
\newcommand{\secref}[1]{\sectionname~\ref{#1}}
\newcommand{\figref}[1]{\figurename~\ref{#1}}
\newenvironment{hoggnumerate}
  {\begin{list}
    {\arabic{enumii}.}
    {\usecounter{enumii}
     %\setlength{\labelwidth}{3em}
     %\setlength{\labelsep}{0em}
     \setlength{\itemsep}{0in}
     \setlength{\parsep}{0in}
     \setlength{\parskip}{0in}
     %\setlength{\leftmargin}{1.5cm}
     %\setlength{\rightmargin}{2cm}
     %\setlength{\itemindent}{0em} 
     %\let\makelabel=\makeboxlabel
    }
  }
{\end{list}}

@@

\begin{abstract}
  Machine learning (ML) methods are having a huge impact across all of the sciences.
  However, ML has a strong ontology---in which only the data exist---and a strong epistemology---in which a model is considered good if it performs well on held-out training data.
  These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
  Here we identify some locations for ML in the natural sciences at which the ontology and epistemology are valuable.
  For example, when an expressive machine learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can make the results more trustworthy.
  We also show that there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
  For one, when ML models are used to emulate physical (or first-principles) simulations, they amplify confirmation biases.
  For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint or ensemble analyses without taking on uncontrolled biases.
  The question in the title is being asked of all of the natural sciences; that is, we are calling on the scientific communities to take a step back and consider the role and value of ML in their fields; the (partial) answers we give here come from the particular perspective of physics.
\end{abstract}
% ] % For the arXiv version, uncomment this square bracket
\printAffiliationsAndNotice{}

@@

This definition is broad, perhaps controversially broad:
In addition to methods like convolutional neural networks \cite{cnn}, multi-layer perceptrons, and transformers \cite{transformer}, it includes large linear regressions, Gaussian processes \cite{gp}, support-vector machines \cite{svm}, principal components analysis \cite{pca}, kernel density estimates \cite{kde}, and even some kinds of multilayer or hierarchical models \cite{multilevel}; indeed almost any models in any current ML textbooks (eg, \citealt{ml_book1, ml_book2}).
Feel free to personally take any more restrictive definition.
Our comments will apply to everything in the larger class.

@@

In Bayesian language, this problem is closely related to the point that when data are to be combined, they should be combined at the likelihood level, not the posterior level.
Likelihoods are not affected by population-level biases.
Likelihoods can produce unbiased point estimates.
Likelihood-based estimators are not usually the lowest-risk estimators, but they can be low- or zero-bias estimators.
The outputs of regressions, on the other hand, are more like posterior-based estimators, affected by the implicit prior set by the training set.
When they are combined, the implicit prior gets amplified.
Most ML regressions are not capable of generating or emulating likelihoods.
There are interesting approaches in development to replace standard regressions with generative models that can produce likelihoods \cite{cannon, likelihood_ratio, sequential, biwei}.

@@

Finally, we note that one objection to this confirmation-bias point is that perhaps it exists just as strongly for any simulation-based inference (such as approximate bayesian computation; \citealt{abc})?
After all, simulations are expensive, so the full suite of them probably can't be affordably repeated.
There are two responses to this objection.
The first is that if the emulators work---that is, if they greatly reduce the cost of the inference---then the emulators have enormously increased the cost difference (or cost factor) between the inference and any re-analysis using physics-based simulations.
The second is that emulators are not doing precisely what the simulations did.
One way to see this is that they are subject to adversarial attacks \cite{melchior}.
Therefore we know that the emulator is not doing precisely what the physics-based simulation is doing.
That is, verification is not just more expensive, but also more important, when an emulator has been introduced.

@@

Given everything we have said here, what are the ways forward for the natural sciences?
ML is critical in many projects; how do we make sure it is doing what we want it to do?
Our view is that the promising directions include imposing physically motivated mathematical structure into the design or regularization of the ML models.
This includes symmetries, conservation laws, and constraints \cite{bronstein2021geometric, villar2023towards}, but also ODE and PDE solvers \cite{karniadakis2021physics}, dynamic programming modules \cite{xu2020can}, and other established mathematical tools.
The research direction that looks towards explainability, interpretability \cite{interpretable}, and trustworthiness \cite{kearns2019ethical} is also fundamental to the development of ML methods for the physical sciences.
These approaches can increase natural scientists’ confidence in ML tools, and also perhaps expand their ontologies and epistemologies.

@@

\paragraph{Acknowledgements:}
It is a pleasure to thank
  Gaby Contardo (SISSA),
  Jennifer Hill (NYU),
  Adrian Price-Whelan (Flatiron),
  Hans-Walter Rix (MPIA),
  Sam Roweis (deceased), and
  Bernhard Sch\"olkopf (MPI-IS)
for conversations over the years related to these arguments, and
  Vedant Chandra (Harvard),
  Contardo,
  Valentino Foit (NYU),
  Alex Jiang (CUNY),
  Charles Margossian (Flatiron),
  Rix,
  Ilya Shpitser (JHU),
  Kate Storey-Fisher (DIPC),
  Jeremias Sulam (JHU),
  Jesse Thaler (MIT),
  Sebastian Wagner-Carena (Flatiron),
  Kaze Wong (Flatiron), and
  our four anonymous ICML reviewers
for valuable comments on this particular manuscript.
SV was partially supported by ONR N00014-22-1-2126, NSF CCF 2212457, the NSF–Simons Research Collaboration
on the Mathematical and Scientific Foundations of Deep Learning (MoDL) (NSF DMS 2031985), and NSF
CAREER 2339682.
The Flatiron Institute is a division of the Simons Foundation.

@@

To illustrate these ideas, we construct a toy data analysis.
This toy will make a general point, but it has a decidedly astronomical feel to it.
The complete toy data set is generated by a process with the algorithm below.
This algorithm makes references to objects $\xi_\eta$ and $\xi_\zeta$ which are generated prior to the start.
\begin{hoggnumerate}
    \item A floating-point value of a known parameter called ``guiding radius'' $r$ is generated in the range $0<r<14$.
    \item A floating-point value of a latent parameter called ``age'' $\eta$ is generated from a Gaussian with mean $14 - r$ and variance 4.
    \item The point is discarded if the age is outside the range $0<\eta<14$. If the point is not discarded:
    \item A $K$-dimensional latent vector $\zeta$ with $K=14$ is generated from a Gaussian with zero mean and variance 1.
    \item A $M$-dimensional latent vector $\xi$ with $M=110$ is created by $\xi = \xi_\eta\,\eta + \xi_\zeta\cdot\zeta$, where $\xi_\eta$ is a $M$-vector and $\xi_\zeta$ is a $M\times K$ matrix.
    The elements of $\xi_\eta$ and $\xi_\zeta$ were drawn (before the start) from Gaussians with zero mean and variances $0.01/K$ and $1/K$ respectively.
    \item A rectified latent vector $\tilde{\xi}$ is created by taking $1 + \xi$ and rectifying all pixels to lie in the range $[0, 1]$.
    \item A $M$-dimensional feature vector $x$ called ``the data'' is generated from a Gaussian with mean $\tilde{\xi}$ and variance $0.0025$. (These feature vectors have been designed to look a little like stellar spectra.)
    \item A label $y$ called ``measured age'' is generated from a Gaussian with mean $\eta$ and variance 1.
\end{hoggnumerate}
The data generated this way are shown in the top panels of \figref{fig:regression}.
The data contain a non-trivial relationship between label $y$ (measured age) and known parameter $r$ (guiding radius).
This relationship is shown in \figref{fig:regression}.
It isn't precisely linear because of the removal of data points with latent ages $\eta$ outside the range $0<\eta<14$.
